\bf No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction

### 2024-10-19 paper.tex
Initial commit

---
title: "No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction"
blinded: 0
authors: 
- name: Adam Bartonicek
  affiliation: Department of Statistics, The University of Auckland
- name: Simon Urbanek
  affiliation: Department of Statistics, The University of Auckland
- name: Paul Murrell
  affiliation: Department of Statistics, The University of Auckland
keywords:
- Interactive data visualization, category theory, abstract algebra, the Gestalt Principles, descriptive statistics
abstract: |
  Interactive data visualization has become a staple of modern data presentation. Yet, despite its growing popularity, there still remains a lack of a formal framework for turning raw data into summary statistics that can then be displayed by interactive graphics. This gap may stem from a subtle yet profound issue: while we would often like treat statistics, graphics, and interaction as independent entities, they are in fact deeply connected. This paper aims to shed light on this interdependence by relating it to two fundamental concepts from category theory: groups and monoids. Specifically, if we want our graphics to support interactive features which break the data into multiple parts and then combine them back together (such as in the case of linked selection), then we need the statistics in our plots to fulfill certain algebraic properties. Ultimately, understanding this hidden structure may empower data visualization experts to build more flexible and expressive interactive data visualization systems.   
bibliography: references.bib
biblio-style: apalike
nocite: |
  @magritte1929
  @horst2020
output:
  rticles::asa_article:
    includes:
      in_header: "preamble.tex"
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{tikz-cd}
---

```{r}
#| echo: false

# knitr::opts_chunk$set(
#   cache = TRUE
# )

```

# Introduction \label{sec:intro}

> "Effective graphical analysis makes things seem obvious, the effort involved in making the graphical analysis effective is not so obvious." [@unwin2018]

The essence of data visualization is comparison. When we visualize, we split our data into parts, summarize each part one or more summary statistics, encode the summaries into visual attributes such as position, size, or color, and finally use these attributes to draw geometric objects [@bertin1983; @wilkinson2012; for a recent review, see @franconeri2021; @wilke2019]. For example, when drawing a typical barplot, we split our data into parts based on the levels of some categorical variable, count the number of cases in each part, and finally represent the counts by drawing bars of the corresponding heights. In this way, each bar becomes a visual representation of its own "small data set", a subset of the original data.

To ground this idea in a more concrete example that will be used throughout this paper, imagine going through your bookshelf, sorting the books by genre, and stacking them on top of each other. You would construct a physical version of the barplot, forming one "bar" (stack) per genre: one bar for sci-fi novels, one bar for romance, one bar for detective fiction, and so on. Even from this very simple example, interesting mathematical properties jump out: for example, a single book cannot go into two stacks (the subsets of the data are *disjoint*) and if all of the stacks are combined back together, we should end up with the bookshelf we started with [the subsets are *surjective* and cover the entirety of the data set, see @ziemkiewicz2009].

Now, just as we can learn about the contents of our library by organizing books in the way we just described, we can learn about our data by splitting it into parts. A natural question then arises: can we learn more by subdividing our data into even smaller parts and comparing these? For instance, in our book example, we can split each stack of books into two smaller stacks based on the gender of the author. Each smaller stack will then represent the unique combination (Cartesian product) of the two categorical variables (genre and gender). Moreover, there arises a hierarchical relationship between the stacks: by combining two genre-gender stacks (child components) together, we can recover the original genre stack (parent component). Similar situation often arises in interactive graphics, where we use techniques such as linked selection to highlight specific subsets of our data. Either way, by forming these finer partitions, we can represent our data with greater granularity and discover trends that may otherwise hide in the aggregate.

However, once we have subdivided each category into its child components, we need to represent the information in a way that visually preserves the hierarchical nature of this relationship. In data visualization, there are two popular methods for doing this, known as "stacking" and "dodging" (see Figure \ref{fig:playfair}). These methods have existed at least since the time of William Playfair [-@playfair1801; -@playfair1822]. Let's briefly illustrate them on the example of the classical barplot. In a stacked barplot, bar segments representing the product of the two factor variables are plotted vertically on top of each other ("stacked"; the operation can be also thought of as "highlighting" parts of the bar). In contrast, in a dodged barplot, bars are plotted side-by-side, as clusters of bars. Finally, there is also a third popular method, known as layering, where we plot bar segments in separate graphical layers and use partial transparency to mitigate overplotting (this technique may have been a bit challenging for William Playfair but has been made simple with computers).

```{r playfair}
#| echo: false
#| fig-align: "center"
#| fig-cap: "Examples of the three methods for dealing with bars that have subdivided into segments. Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010)."
#| message: false
#| warning: false

library(ggplot2)
library(patchwork)
library(magick)

p1 <- image_read("./images/playfair1.jpg") |> image_ggplot()
p2 <- image_read("./images/playfair2.jpg") |> image_ggplot()

mpg$manufacturer <- tools::toTitleCase(mpg$manufacturer)

p3 <- ggplot(mpg, aes(manufacturer, fill = factor(year))) +
  geom_bar(position = "identity", alpha = 0.5) +
  scale_x_discrete(labels = abbreviate) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
)

(p1 + p2) / p3
```

Much has been written about the relative merits of stacking, dodging, and layering. For example, layering is only useful with few categories, as blending many colors can make it difficult to tell the categories apart [@franconeri2021; @wilke2019]. Further, in a landmark study, @cleveland1984 showed that people tend to be less accurate when reading information from stacked bar charts as opposed to dodged bar charts. Specifically, since the lower y-axis coordinate of a stacked segment is pushed up by the cumulative height of the segments below it, it becomes difficult to accurately compare segments' length, both within and across bars [@cleveland1984]. Subsequent research has independently validated these findings and expanded upon them [see e.g. @heer2010; @thudt2016; @quadri2021]. Ultimately, the suboptimal statistical legibility of stacking has led some authors to advise against its use altogether [@kosara2016; @wilke2019], or at least urge caution [see e.g. @byron2008; @cairo2014; @franconeri2021].

However, despite the mixed reception towards stacking in static visualization, it remains a popular choice in interactive data visualization. Among popular interactive data visualization libraries, stacked plots such as bar charts and histograms make frequent appearance [see e.g. @cook2023; @murray2017; @vegalite2022; @vanderplas2018; @sievert2020; @swayne2003; @xie2014; @theus2002; @urbanek2003; @urbanek2011; @wills2008; @ware2019]. What is the reason for this? Has the interactive data visualization community simply failed to catch up with the current best practices? Or is there perhaps more going on?

We argue that not only is stacking a useful technique, but that it can in fact offer valuable insights about the mathematical structure underlying our visualizations. To make the strongest case possible, we bring forward evidence from several levels of the visualization process. We start by discussing the properties of visual perception and how they apply to both static and interactive visualizations. Next, we explore the process of computing statistical summaries that form the basis of our plots. Finally, we dive into the algebraic properties of the functions underlying these statistical summaries. Throughout the text, we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and efficient interactive graphics, we simply cannot treat graphics, statistics, and interaction as independent components. Instead we need to consider them holistically, within the context of their respective visual and mathematical properties. Understanding this deep web of interdependence may in turn empower us to innovate and build new kinds of robust and expressive interactive visualizations.

# Visual Perception and Interactivity

## Visual perception and the Gestalt Principles

Over the course of our evolutionary history, the human brain has evolved a sophisticated pipeline for processing visual information. This pipeline has been the subject of a great body of research in cognitive neuroscience [@goebel2004; @knudsen2020; for a brief overview, see @ware2019, pp. 20]. Many of the findings from this research are directly applicable to data visualization. For example, certain particularly salient visual stimuli are perceived automatically, rapidly (<10 ms), in parallel, and at a constant speed regardless of the number of distractors [@treisman1985]. This phenomenon, known as pre-attentive processing, is key for data visualization, as the right use of visual attributes such as color, contrast, or shape can help us design figures which communicate insights much more effectively [@ware2019].

However, many of the findings from visual perception research are quite specific; to guide effective visualization, we also need some general, high-level rules. One such set of high-level rules are the Gestalt principles of visual perception [see e.g. @cairo2012; @ware2019; @rosli2015; @vanderplas2020; @todorovic2008; @pinker1990]. These principles formulate a set of fundamental "laws" related to the perception of patterns, primarily visual. They were discovered by the founders of the Gestalt school of psychology early in the 20th century, primarily by means of self-observation, and, despite not being grounded in sound understanding of the neural mechanisms of the brain, have nevertheless been later independently supported by findings from cognitive neuroscience [@guberman2017; @ware2019; @wertheimer1938].

Among the Gestalt principles, there are a few which are particularly relevant to data visualization. First, there are the two principles of proximity and similarity. These principles state that objects that which appear in close together and/or similar in appearance tend to be perceived as part of single group. Second, the closure and common region principles state that regions with closed contours tend to be perceived as single, unified objects. Finally, the figure and ground principle states we tend to see one part of a scene as the central object (the "figure") while the rest is perceived as the background (the "ground"). This can be modulated by other Gestalt principle: for example, smaller, darker, and symmetric objects with closed contours are more likely to be perceived as the figure [@ware2019].

Whether we intend them to or not, the Gestalt principles affect how our data visualizations are perceived. As such, it is a good idea to try to use them to our advantage. Sometimes, little effort is needed. Such is the case, for example, when identifying clusters in a scatterplot [@rosli2015]: the principles of proximity and similarity ensure that clouds of similarly shaped and colored points will be perceived as discrete groups by the observer. In other cases, however, our design choices can have a significant impact. For instance, another way to have objects to be perceived as part of a group is to draw a closed shape around them [@cairo2012]. Alternatively, if we want to highlight a particular part of a data set, we can leverage the figure and ground principle and use attributes such as color and saturation to make it stand out [for specific examples, see @franconeri2021; @wilke2019, p. 34].

Finally, ignoring or violating the Gestalt principles can make our graphics less legible or even misleading. @tufte2001 had documented many such violations, such as deceptive 3D effects, inadvertent optical illusions caused by dense stripe patterns, and the overuse of superfluous graphical elements he termed "chartjunk". Similarly, @cairo2019 highlights many cases of graphics that violate the Gestalt principles. Finally, there is also the issue of well-designed graphics that communicate no statistical information at all, or even information that is blatantly false [@cairo2014; @franconeri2021; @cairo2019; @gelman2013; @tufte2001]. Whether intentional or not, poor design can have a significant impact on the legibility and effectiveness of statistical graphics.

## Gestalt and Interactivity

The Gestalt principles apply to both static and interactive visualizations, however, the way they apply is different. Static data visualizations can be designed well or poorly, but this design quality does not change over time. In contrast, interactive visualizations change over time, with user interaction. As such, an interactive visualization may conform to the Gestalt principles at one instance of time and violate them in another. In fact, interactive visualizations can even differ in how well they leverage design principles throughout time: the animations that interactions induce may themselves conform to good or poor design [see @hullman2013; @ware2019].

@@

## Case study: Linked Selection

Linked selection, also known as linked brushing or linked highlighting, has been consistently ranked as one of the most useful interactive features in data visualization [see e.g. @buja1996; @heer2012; @wilhelm2003; @ware2019; @ward2015]. By clicking or clicking-and-dragging over objects in one plot, users can highlight parts of objects corresponding to the selected cases across all other "linked" plots. This allows the user to rapidly "drill-down" [@dix1998; @theus2002] and explore trends across different dynamically-generated subsets of the data [similar to interactive filtering, see e.g. @heer2012]. The ability to quickly materialize different views of the data makes linked selection a versatile tool for data exploration.

Yet, despite its usefulness, linked selection also imposes some fundamental constraints on the figure. Specifically, all objects we visualize need to be able to:

1. **Be selected**
2. **Represent selection**
3. (both in a visually and statistically sound way).

While these constraints may not seem too restrictive, we argue that they pose significant challenges for certain types of plots and visualization styles. Specifically, some kinds of plots may produce undesirable visual behavior when combined with linked selection. In fact, this may explain why stacking is preferred over dodging within the interactive data visualization community. Let's explore this idea in more depth by comparing linked selection with a stacked vs. dodged barplot.

```{r display}
#| echo: false
#| fig-height: 3.5
#| fig-cap: "\\label{fig:display}The visual coherence of stacking vs. dodging
#| during linked selection on the `mtcars` data set [@henderson1981]. Bars show
#| the numbers of cars in the 4-, 6-, and 8-cylinder categories. 
#| Dark segments represent selected cases, light segments represent unselected cases, 
#| and the plots' contour is shown via a light outline. Plots left to right show more cars being selected. 
#| Top row: the contour of a stacked barplot remains constant throughout selection
#| (three bars of the same height). Bottom row: the contour of a dodged barplot 
#| changes dramatically with selection. Note the changes to the upper y-axis limit 
#| and the absence of the lighter segment among the 6-cylinder cars
#| in the rightmost plot (since all 6-cylinder cars are selected, 
#| the darker segment has been stretched to fill the available space; 
#| `ggplot2` default for handling empty subcategories)."

library(ggplot2)

cols <- c('grey80', '#56B4E9', '#009E73')
mtcars$cyl <- factor(mtcars$cyl)
plots <- list()

p0 <- ggplot() +
  guides(fill = "none") +
  scale_fill_manual(values = c(cols[1], cols[3])) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_text(size = 12),
  )

set.seed(89719)
mtcars$group1 <- factor(sample(rep(c(0, 1), c(27, 5))))
mtcars$group2 <- factor(sample(rep(c(0, 1), c(17, 15))))
mtcars$group3 <- factor(sample(rep(c(0, 1), c(7, 25))))

for (i in 1:3) {
  mtcars$group <- mtcars[[paste0("group", i)]]
  plots[[i]] <- p0 + 
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
             col = cols[2]) +
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group)) +
    labs(title = ifelse(i == 1, "Stacking", ""))
  plots[[3 + i]] <- plot_spacer()
  plots[[6 + i]] <- p0 + 
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
             position = "dodge", col = cols[2]) +
    geom_bar(data = mtcars, mapping = aes(cyl, fill = group), 
             position = "dodge") +
    labs(title = ifelse(i == 1, "Dodging", ""))
}

wrap_plots(plots, nrow = 3, heights = c(1, 0.5, 1))
```

First, we propose that stacking is the superior method for *displaying* selection (see Figure \ref{fig:display}). In a stacked barplot, each stacked bar presents itself as a single visual entity, thanks to the four Gestalt principles: proximity, similarity, closure, and common region. Since the stacked segments are placed right on top of each other, and share the same width and closed border, they are perceived as part of a unified whole (the bar). This remains true throughout selection - no matter how the heights of the highlighted segments within a stacked bar change, the height of the whole bar remains constant, and so does the overall outline of the plot.

Dodged barplots lack this degree of visual consistency. In a dodged barplot, bar segments placed side-by-side in clusters, and, since each segment can be of different length, these clusters lack a shared border (common region). This leads to several design challenges. First, when selection occurs, the jagged contour of the dodged barplot can change dramatically as more cases are added or removed. This creates distracting visual clutter that makes it difficult to track changes over time. Second, since the heights of the bar segments are not bound by any value present in the plot (unlike the heights of the segments in the stacked barplot, which are bound by the height of the tallest bar), we need to decide how the handle the upper y-axis limit in response to selection. One option is to make the upper y-axis limit dynamic. If we do this, we ensure that the bar segments always fit the plotting region, however, we also lose the context that the upper y-axis limit provides each time selection occurs. Alternatively, we can assign a static value to the upper y-axis limit, however, then we risk the bar segments growing outside of the bounds of the plotting region.

```{r selection, eval=FALSE}
#| echo: false
#| fig-height: 2.5
#| fig-cap: "\\label{fig:selection}Selection behavior in stacked vs. dodged barplot. Darker segments represent selected cases, dashed rectangle represent selection region created by clicking and dragging.
#| Left: in a stacked barplot, it is natural to select an entire category by clicking/dragging over a portion of a bar, regardless of the selection status, e.g. here the dashed selection rectangle would select the entire category A.
#| Right: in a dodged barplot, it is not clear what to expect from clicking and dragging over a segment - is the intention to select only the cases corresponding to the segment, or the cases corresponding to the whole bar? How does the plot change when the selection happens?"

library(ggplot2)
library(patchwork)

mtcars$am <- factor(mtcars$am)
mtcars$cyl <- factor(mtcars$cyl, labels = c("A", "B", "C"))

p0 <- ggplot() +
  scale_fill_manual(values = c("grey80", "indianred")) +
  guides(fill = "none") +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
)

p1 <- p0 + 
  geom_bar(data = mtcars, mapping = aes(cyl, fill = am)) +
  annotate("rect", xmin = 0.5, xmax = 1.25, ymin = 9.5, ymax = 11.5,
           fill = NA, col = "grey20", lty = "dashed", linewidth = 0.4)

p2 <- p0 + geom_bar(data = mtcars, mapping = aes(cyl, fill = am), 
                    position = "dodge") +
  annotate("rect", xmin = 0.75, xmax = 1.25, ymin = 7, ymax = 8.5,
           fill = NA, col = "grey20", lty = "dashed", linewidth = 0.4)

p1 + p2
```

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage[]{natbib}
\usepackage{textcomp}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

%% load any required packages here

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amssymb}
\usepackage{tikz-cd}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\hypersetup{
  pdftitle={No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction},
  pdfkeywords={Interactive data visualization, category theory, abstract
algebra, the Gestalt Principles, descriptive statistics},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf No More, No Less Than Sum of its Parts: The Algebra of
Graphics, Statistics, and Interaction}

  \author{
        Adam Bartonicek \\
    Department of Statistics, The University of Auckland\\
     and \\     Simon Urbanek \\
    Department of Statistics, The University of Auckland\\
     and \\     Paul Murrell \\
    Department of Statistics, The University of Auckland\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf No More, No Less Than Sum of its Parts: The Algebra of
Graphics, Statistics, and Interaction}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Interactive data visualization has become a staple of modern data
presentation. Yet, despite its growing popularity, there still remains a
lack of a formal framework for turning raw data into summary statistics
that can then be displayed by interactive graphics. This gap may stem
from a subtle yet profound issue: while we would often like treat
statistics, graphics, and interaction as independent entities, they are
in fact deeply connected. This paper aims to shed light on this
interdependence by relating it to two fundamental concepts from category
theory: groups and monoids. Specifically, if we want our graphics to
support interactive features which break the data into multiple parts
and then combine them back together (such as in the case of linked
selection), then we need the statistics in our plots to fulfill certain
algebraic properties. Ultimately, understanding this hidden structure
may empower data visualization experts to build more flexible and
expressive interactive data visualization systems.
\end{abstract}

\noindent%
{\it Keywords:} Interactive data visualization, category theory,
abstract algebra, the Gestalt Principles, descriptive statistics

\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!

\section{\texorpdfstring{Introduction
\label{sec:intro}}{Introduction }}\label{introduction}

\begin{quote}
``Effective graphical analysis makes things seem obvious, the effort
involved in making the graphical analysis effective is not so obvious.''
\citep{unwin2018}
\end{quote}

The essence of data visualization is comparison. When we visualize, we
split our data into parts, summarize each part by a set of one or more
statistics, and then encode these summaries into visual channels such as
position, size, or color that we then use to draw geometric objects
\citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
example, when drawing a typical barplot, we split our data based on the
levels of some categorical variable, count or sum up the values of some
other variable, and finally represent these summaries by drawing bars of
the corresponding heights. In this way, each bar becomes a visual
representation of its own ``small data set'', a subset of the original
data.

To ground this idea in a more concrete example, one that will be used
throughout the rest of this paper, imagine going through your bookshelf,
sorting the books by genre, and stacking them on top of each other. You
would construct a physical version of the barplot, forming one ``bar''
(stack) per genre: one bar for sci-fi novels, one bar for romance, one
bar for detective fiction, and so on. Even from this very simple
example, interesting mathematical properties jump out: for example, a
single book cannot go into two stacks (the subsets of the data are
\emph{disjoint}) and if all of the stacks are combined back together, we
should end up with the bookshelf we started with \citep[the subsets are
\emph{surjective} and cover the entirety of the data set,
see][]{ziemkiewicz2009}.

Now, just as we can learn about the contents of our library by
organizing books in the way we just described, we can learn about our
data by splitting it into parts. A natural question that arises is: can
we learn more by subdividing our data into even smaller parts and
comparing these? For instance, in our book example, we could split each
stack of books into two smaller stacks based on the gender of the
author. Each smaller stack will then represent the unique combination
(Cartesian product) of the two categorical variables (genre-gender).
Moreover, there arises a hierarchical relationship between the stacks:
by combining two genre-gender stacks (child components) together, we can
recover the original genre stack (parent component). Similar situation
often arises in interactive graphics, where we use techniques such as
linked selection to highlight specific subsets of our data. Either way,
by forming these finer partitions, we can represent our data with
greater granularity and discover trends that may otherwise hide in the
aggregate.

However, once we have subdivided each category into its child
components, we need a way of representing this information that visually
preserves the hierarchical nature of the relationship. In data
visualization, there are two popular methods for doing this, known as
``stacking'' and ``dodging'' (see Figure \ref{fig:playfair}). These
methods have existed at least since the time of William Playfair
\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
briefly illustrate them on the example of the classical barplot. In a
stacked barplot, bar segments representing the product of the two factor
variables are plotted vertically on top of each other (``stacked''; the
operation can be also thought of as ``highlighting'' parts of the bar).
In contrast, in a dodged barplot, bars are plotted side-by-side, as
clusters of bars. Finally, there is also a third popular method, known
as layering, where we plot bar segments in separate graphical layers and
use partial transparency to mitigate overplotting (this technique may
have been a bit challenging for William Playfair but has been made
simple with computers).

\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/playfair-1}

}

\caption{Examples of the three methods for dealing with bars that have subdivided into segments. Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010).}\label{fig:playfair}
\end{figure}

Much has been written about the relative merits of stacking, dodging,
and layering. For example, layering is only useful with few categories,
as blending many colors can make it difficult to tell the categories
apart \citep{franconeri2021, wilke2019}. Further, in a landmark study,
\citet{cleveland1984} showed that people tend to be less accurate when
reading information from stacked bar charts as opposed to dodged bar
charts. Specifically, since the lower y-axis coordinate of a stacked
segment is pushed up by the cumulative height of the segments below it,
it becomes difficult to accurately compare segments' length, both within
and across bars \citep{cleveland1984}. Subsequent research has
independently validated and expanded upon these findings \citep[see
e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
statistical legibility of stacking has led some authors to advise
against its use altogether \citep{kosara2016, wilke2019}, or at least
urge caution \citep[see e.g.][]{byron2008, cairo2014, franconeri2021}.

However, despite the mixed reception of stacked plots in static
visualization, they remain a popular choice in interactive data
visualization. Among popular interactive data visualization libraries,
stacked plots such as bar charts and histograms make frequent appearance
\citep[see
e.g.][]{cook2023, murray2017, vegalite2022, vanderplas2018, sievert2020, swayne2003, xie2014, theus2002, urbanek2003, urbanek2011, wills2008, ware2019}.
What is the reason for this? Has the interactive data visualization
community simply failed to catch up with the current best practices? Or
is there perhaps more going on?

We argue that not only is stacking a useful technique, but that it can
in fact offer valuable insights about the mathematical structure
underlying our visualizations. To make the strongest case possible, we
bring forward evidence from several levels of the visualization process.
We start by discussing the properties of visual perception and how they
apply to both static and interactive data visualizations. Next, we
explore the process of computing statistical summaries that form the
basis of our plots. Finally, we dive into the algebraic properties of
the functions underlying these statistical summaries. Throughout the
text, we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and
efficient interactive graphics, we simply cannot treat graphics,
statistics, and interaction as independent components. Instead we need
to consider them holistically, within the context of their respective
visual and mathematical properties. Understanding this deep web of
interdependence may in turn empower us to innovate and build new kinds
of robust interactive visualizations.

\section{Visual Perception and
Interactivity}\label{visual-perception-and-interactivity}

\subsection{Visual perception and the Gestalt
Principles}\label{visual-perception-and-the-gestalt-principles}

Over the course of our evolutionary history, the human brain has evolved
a sophisticated pipeline for processing visual information. This
pipeline has been the subject of a great body of research in cognitive
neuroscience
\citetext{\citealp{goebel2004}; \citealp{knudsen2020}; \citealp[for a
brief overview, see][20]{ware2019}}. Many of the findings from this
research are directly applicable to data visualization. For example,
certain particularly salient visual stimuli are perceived automatically,
rapidly (\textless10 ms), in parallel, and at a constant speed
regardless of the number of distractors \citep{treisman1985}. This
phenomenon, known as pre-attentive processing, is key for data
visualization, as the right use of visual attributes such as color,
contrast, or shape can help us design figures which communicate insights
much more effectively \citep{ware2019}.

However, many of the findings from visual perception research are quite
specific; to guide effective visualization, we also need some general,
high-level rules. One such set of high-level rules are the Gestalt
principles of visual perception \citep[see
e.g.][]{cairo2012, ware2019, rosli2015, vanderplas2020, todorovic2008, pinker1990}.
These principles formulate a set of fundamental ``laws'' related to the
perception of patterns, primarily visual. They were discovered by the
founders of the Gestalt school of psychology early in the 20th century,
primarily by means of self-observation, and, despite not being grounded
in sound understanding of the neural mechanisms of the brain, have
nevertheless been later independently supported by findings from
cognitive neuroscience \citep{guberman2017, ware2019, wertheimer1938}.

Among the Gestalt principles, there are a few which are particularly
relevant to data visualization. First, there are the two principles of
proximity and similarity. These principles state that objects that which
appear in close together and/or similar in appearance tend to be
perceived as part of single group. Second, the closure and common region
principles state that regions with closed contours tend to be perceived
as single, unified objects. Finally, the figure and ground principle
states we tend to see one part of a scene as the central object (the
``figure'') while the rest is perceived as the background (the
``ground''). This can be modulated by other Gestalt principle: for
example, smaller, darker, and symmetric objects with closed contours are
more likely to be perceived as the figure \citep{ware2019}.

Whether we intend them to or not, the Gestalt principles how we perceive
data visualizations. As such, it is a good idea to try to use them to
our advantage. Sometimes, little effort is needed. Such is the case, for
example, when identifying clusters in a scatterplot \citep{rosli2015}:
the principles of proximity and similarity ensure that clouds of
similarly shaped and colored points will be perceived as discrete groups
by the observer. In other cases, however, our design choices can have a
significant impact. For instance, another way to have objects to be
perceived as part of a group is to draw a closed shape around them
\citep{cairo2012}. Alternatively, if we want to highlight a particular
part of a data set, we can leverage the figure and ground principle and
use attributes such as color and saturation to make it stand out
\citep[for specific examples, see][34]{franconeri2021, wilke2019}.

Finally, ignoring or violating the Gestalt principles can make our
graphics less legible or even misleading. \citet{tufte2001} had
documented many such violations, such as deceptive 3D effects,
inadvertent optical illusions caused by dense stripe patterns, and the
overuse of superfluous graphical elements he termed ``chartjunk''.
Similarly, \citet{cairo2019} highlights many cases of graphics that
violate the Gestalt principles. Finally, there is also the issue of
well-designed graphics that communicate no statistical information at
all, or even information that is blatantly false
\citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
Either way, whether intentional or not, poor design can have a
significant impact on the legibility and effectiveness of statistical
graphics.

\subsection{Gestalt and Interactivity}\label{gestalt-and-interactivity}

The Gestalt principles apply to both static and interactive
visualizations, however, the way they apply is different. Static data
visualizations can be designed well or poorly, but this design quality
does not change over time. In contrast, interactive visualizations
change over time, with user interaction. As such, an interactive
visualization may conform to the Gestalt principles at one instance of
time and violate them in another. In fact, interactive visualizations
can even differ in how well they leverage design principles throughout
time: the animations that interactions induce may themselves conform to
good or poor design \citep[see][]{hullman2013, ware2019}.

@@

\subsection{Case study: Linked
Selection}\label{case-study-linked-selection}

Linked selection, also known as linked brushing or linked highlighting,
has been consistently ranked as one of the most useful interactive
features in data visualization \citep[see
e.g.][]{buja1996, heer2012, wilhelm2003, ware2019, ward2015}. By
clicking or clicking-and-dragging over objects in one plot, users can
highlight parts of objects corresponding to the selected cases across
all other ``linked'' plots. This allows the user to rapidly
``drill-down'' \citep{dix1998, theus2002} and explore trends across
different dynamically-generated subsets of the data \citep[similar to
interactive filtering, see e.g.][]{heer2012}. The ability to rapidly
materialize different views of the data makes linked selection a
particularly versatile tool for data exploration.

Yet, despite its versatility, linked selection also imposes some
fundamental constraints on the visualization. Specifically, all objects
we visualize need to be able to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Be selected}
\item
  \textbf{Represent selection}
\item
  (both in a visually and statistically sound way).
\end{enumerate}

While these constraints may not seem too restrictive, we argue that they
pose significant challenges for certain types of plots and visualization
styles. Specifically, some kinds of plots may produce undesirable visual
behavior when combined with linked selection. In fact, this may explain
why stacking is preferred over dodging within the interactive data
visualization community. Let's explore this idea in more depth by
comparing linked selection with a stacked vs.~dodged barplot.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/display-1.pdf}
\caption{\label{fig:display}The visual coherence of stacking vs.~dodging
during linked selection on the \texttt{mtcars} data set
\citep{henderson1981}. Bars show the numbers of cars in the 4-, 6-, and
8-cylinder categories. Dark segments represent selected cases, light
segments represent unselected cases, and the plots' contour is shown via
a light outline. Plots left to right show more cars being selected. Top
row: the contour of a stacked barplot remains constant throughout
selection (three bars of the same height). Bottom row: the contour of a
dodged barplot changes dramatically with selection. Note the changes to
the upper y-axis limit and the absence of the lighter segment among the
6-cylinder cars in the rightmost plot (since all 6-cylinder cars are
selected, the darker segment has been stretched to fill the available
space; \texttt{ggplot2} default for handling empty subcategories).}
\end{figure}

First, we propose that stacking is the superior method for
\emph{displaying} selection (see Figure \ref{fig:display}). In a stacked
barplot, each stacked bar presents itself as a single visual entity,
thanks to the four Gestalt principles: proximity, similarity, closure,
and common region. Since the stacked segments are placed right on top of
each other, and share the same width and closed border, they are
perceived as part of a unified whole (the bar). This remains true
throughout selection - no matter how the heights of the highlighted
segments within a stacked bar change, the height of the whole bar
remains constant, and so does the overall outline of the plot.

Dodged barplots lack the degree of visual consistency found in stacked
barplots. In a dodged barplot, the clusters of segments placed
side-by-side lack a shared border (common region), leading to several
design challenges. First, when selection occurs, the jagged contour of
the dodged barplot can change dramatically as more cases are added or
removed. This creates distracting visual clutter that makes it difficult
to track changes over time. Second, selection may alter the height of
the tallest bar segment in the plot. Since the height of the bar
segments is not bound by the height of the tallest bar (as is the case
in stacked barplot), this presents us with a dilemma. Either either we
make the upper y-axis limit dynamic, meaning that the bar segments will
always fit the plotting region, but then the context that the upper
y-axis limit provides is lost with each time selection happens.
Alternatively, we keep the upper y-axis limit static, but then we risk
the bar segments growing outside of the bounds of the plotting region.

Stacked barplots offer several other advantages over dodged bars within
the context of linked selection. First, we propose that stacking also
provides a simpler mental model for \emph{selection}. Since each stacked
bar is a single visual entity, it makes intuitive sense that clicking it
will select the entire category corresponding to it. Indeed, this is how
many interactive data visualization systems implement linked selection:
clicking on part of an object selects the category corresponding to the
entire object \citep[see
e.g.][]{cook2023, vegalite2022, vanderplas2018, xie2014, theus2002, urbanek2003, urbanek2011, swayne2003}.
In dodged barplots, on the other hand, the expectation is less clear:
when we click on a bar segment in a dodged barplot, will we select the
entire category or just the cases corresponding to the segment? Second,
in a stacked barplot, the bar width and inter-bar gap are constant
throughout selection. In a dodged barplot, it is not clear how to handle
the gaps when some selection groups are empty. For example, when no
cases are selected, should we leave gaps for the (currently absent)
highlighted segments, or draw wide bars and only shrink these in
response to selection? Finally, by presenting fewer visual entities on
the screen, stacked bars may also reduce cognitive load
\citetext{\citealp{sweller2019}; \citealp[for a non-technical overview,
see][]{knaflic2015}}.
